---
title: "exercise solutions"
format:
  html:
    embed-resources: true
editor_options: 
  chunk_output_type: console
---

There are many ways to achieve the same thing in R. This answer key provides one or more solutions aligned with the workshop materials, but there are other ways to successfully complete the exercises.

# Data Analysis

## Exercise: Import the Data

Copy the `gov_env_2015.csv` file from the workshop materials folder to the folder for the new RStudio project you created for the exercises.

Add code to your script to import the data. Make sure to assign the data a name.

Write code to get the dimensions of the data frame and the names of the variables (columns).

```{r}
# the path to the file may differ for you
govenv <- read.csv("data/gov_env_2015.csv")

dim(govenv)
nrow(govenv)
ncol(govenv)

names(govenv)
```

## Exercise: Load external packages

Write code to load the `tidyverse` packages into your R session.

```{r}
library(tidyverse)
```

## Exercise: Compute Summary Statistics

Compute the average (mean) and median values of `globalization`, `rain_avg`, `temp_avg`, and `co2_per_capita`.

```{r}
# with base R 
mean(govenv$globalization)
mean(govenv$rain_avg)
mean(govenv$temp_avg)
mean(govenv$co2_per_capita, na.rm=TRUE)

median(govenv$globalization)
median(govenv$rain_avg)
median(govenv$temp_avg)
median(govenv$co2_per_capita, na.rm=TRUE)

# with tidyverse (simple but doesn't ignore NA values)
govenv %>% 
  summarise(across( c(globalization, rain_avg, temp_avg, co2_per_capita),
                    list( mean = mean, median = median)))

# with tidyverse - advanced, can ignore NA values using anonymous functions
govenv %>% 
  summarise(across( c(globalization, rain_avg, temp_avg, co2_per_capita),
                    list( mean = ~mean(.x, na.rm = TRUE), 
                          median = ~median(.x, na.rm = TRUE))))

?across # check documentation for more details
```

## Exercise: Frequency Table

Create a frequency table of the number of natural disasters (`nat_disaster_n`).

Create a frequency table of the number of climate laws/policies (`climate_laws_policies`).

Be sure to include any missing values in the tables.

```{r}
table(govenv$nat_disaster_n, useNA="ifany")
table(govenv$climate_laws_policies, useNA="ifany")
```

## Exercise: Conditional Summary Statistics

Compute the average (mean) of `rain_avg` and `temp_avg` for countries in the region "Eastern Europe/Former USSR".

Challenge: Which country in "Latin America"" had the highest value of `co2_per_capita`?

Challenge: How many countries in "Sub-Saharan Africa" had `co2_per_capita` greater than 1 metric ton?

```{r}
# base R
mean(govenv$rain_avg[govenv$region == "Eastern Europe/Former USSR"])
mean(govenv$temp_avg[govenv$region == "Eastern Europe/Former USSR"])

maxval <- max(govenv$co2_per_capita[govenv$region == "Latin America"])
govenv$country[govenv$co2_per_capita == maxval]

sum(govenv$co2_per_capita[govenv$region == "Sub-Saharan Africa"] > 1, na.rm=TRUE)


# tidyverse
govenv %>% 
  filter(region == "Eastern Europe/Former USSR") %>% 
  summarise(across(c(rain_avg, temp_avg), list(mean = mean)))

govenv %>% 
  filter(region=="Latin America") %>% 
  filter(co2_per_capita == max(co2_per_capita, na.rm=T)) %>% 
  select(country)

govenv %>% 
  filter(region=="Sub-Saharan Africa", co2_per_capita > 1) %>% 
  nrow()
```

## Exercise: Count and Find Missing Data

Which variable(s) have the most missing values?

Challenge: Are democracies or not democracies missing more `co2_per_capita` values?

Challenge: Which countries are missing `co2_growth_rate` values?

```{r}
colSums(is.na(govenv))

# Challenge
table(govenv$democracy, is.na(govenv$co2_per_capita))
# alternative
sum(is.na(govenv$co2_per_capita[govenv$democracy == 1]))
sum(is.na(govenv$co2_per_capita[govenv$democracy == 0]))

# Challenge
govenv$country[is.na(govenv$co2_growth_rate)]
```

## Exercise: Subset the Data Frame: Compound Conditions

Subset the data frame to only include monarchies in "Western Europe, North America, Australia, New Zealand". Save the subset in a new variable. Which of the countries in the subset has the largest population?

```{r}
# base R
west_countries <- govenv[govenv$region == "Western Europe, North America, Australia, New Zealand" & govenv$monarchy == 1,]
west_countries$country[west_countries$population == max(west_countries$population)]

# tidyverse
west_countries <- filter(govenv, region == "Western Europe, North America, Australia, New Zealand", monarchy == 1)
filter(west_countries, population == max(population)) %>%
  select(country)
```

## Exercise: Modify Variables

Modify the `monarchy` variable so that the values are "monarchy" and "not monarchy" instead of numeric or logical values.

Convert `rain_avg` to inches instead of mm (multiply by 25.4).

Change country name of "Congo, Democratic Republic" to "DR Congo".

```{r}
govenv$monarchy <- ifelse(govenv$monarchy == 1, "monarchy", "not monarchy")
# alternative
govenv <- mutate(govenv, monarchy = ifelse(monarchy == 1, "monarchy", "not monarchy"))

govenv$rain_avg <- govenv$rain_avg * 25.4
# alternative
govenv <- mutate(govenv, rain_avg * 25.4)

govenv$country[govenv$country == "Congo, Democratic Republic"] <- "DR Congo"
# alternative
govenv$country <- ifelse(govenv$country == "Congo, Democratic Republic", "DR Congo", govenv$country)
# alternative
govenv <- mutate(govenv, country = ifelse(country == "Congo, Democratic Republic", "DR Congo", country))
```

## Exercise: Change Variable Names

Rename the `ag_land_pct` variable to `agriculture_land_pct`.

```{r}
govenv <- rename(govenv, agriculture_land_pct = ag_land_pct)

# alternative
names(govenv)[names(govenv) == "ag_land_pct"] <- "agriculture_land_pct"

# alternative (above options are more robust to changes in the dataset)
names(govenv) ## check to find position
names(govenv)[15] <- "agriculture_land_pct"
```

## Exercise: Modify Variable Types

Change the presidential variable to be of type logical (TRUE/FALSE) instead of numeric.

```{r}
govenv$presidential <- as.logical(govenv$presidential)

# alternative
govenv <- mutate(govenv, presidential = as.logical(presidential))
```

## Exercise: Make a New Variable

Create a new variable in the data frame (make up a name) that is TRUE if any climate laws or policies were passed and FALSE otherwise.

```{r}
govenv$laws_passed <- govenv$climate_laws_policies > 0
```

## Exercise: Create a Factor

Modify the presidential variable to have the values "Presidential" and "Not Presidential". Convert the variable to a factor, where the "Presidential" category displays first.

Challenge: Research (or ask a GenAI tool) how to use labels when creating a factor. Create a factor of `electoral_comp`, adding text labels for the categories.

```{r}
# if presidential is logical (TRUE/FALSE)
govenv$presidential <- factor(ifelse(govenv$presidential, "Presidential", "Not Presidential"),
                              levels = c("Presidential", "Not Presidential"))

# if presidential is numeric (0/1)
# govenv$presidential <- factor(ifelse(govenv$presidential == 1, "Presidential", "Not Presidential"),
#                               levels = c("Presidential", "Not Presidential"))

# Challenge
govenv$electoral_comp <- factor(govenv$electoral_comp,
                                labels=c("No elections", "Single-party elections",
                                         "Non-democratic multi-party elections",
                                         "Democratic elections"))
```

## Exercise: Group Statistics (with dplyr)

Compute the average and maximum of globalization by whether the country is a democracy.

How many deaths were caused by natural disasters by region? (hint: use sum())

Challenge: Compute the mean number of climate_laws_policies grouping countries by whether they have above or below average `co2_per_capita`.

```{r}
govenv %>%
  group_by(democracy) %>%
  summarize(min_global = mean(globalization),
            max_global = max(globalization))

govenv %>%
  group_by(region) %>%
  summarize(nat_disaster_deaths = sum(nat_disaster_deaths, na.rm=TRUE))

govenv %>%
  mutate(high_co2 = co2_per_capita > mean(co2_per_capita, na.rm=TRUE)) %>%
  group_by(high_co2) %>%
  summarize(mean_laws = mean(climate_laws_policies, na.rm=TRUE))
```

## Exercise: Group Proportions

What proportion of countries are democracies?

How many and what proportion of countries fall into each category of electoral_comp?

Challenge: compute the proportion of countries that are democracies per region.

```{r}
govenv %>%
  count(democracy) %>%
  mutate(prop = n / sum(n))
# alternative
library(janitor)
tabyl(govenv, democracy)
# alternative
sum(govenv$democracy == 1)/nrow(govenv)
# alternative
prop.table(table(govenv$democracy))

govenv %>%
  count(electoral_comp) %>%
  mutate(prop = n / sum(n))
# alternative
tabyl(govenv, electoral_comp)
# alternative
table(govenv$electoral_comp)
prop.table(table(govenv$electoral_comp))

# Challenge
govenv %>%
  group_by(region) %>%
  summarize(prop_democracy = sum(democracy == 1)/n())
```

## Exercise: Test Your Reproducibility

Save your code file, and then close RStudio. If prompted as to whether to save the workspace image, say no. Then restart RStudio (if you created an RStudio project, make sure it's open) and re-run your code file through in order to see if everything still works.
