---
title: "Week 7 Exercises (Answers)"
format: html
project:
  execute-dir: project
editor_options: 
  chunk_output_type: console
---


```{r}
library(tidyverse)
library(janitor)
```

Load `data/diabetes.csv`.  Basic information on the dataset is available here: https://hbiostat.org/data/repo/diabetes


```{r}
diabetes <- read_csv("data/diabetes.csv")
```


## Exercise 

Is the proportion of women in the data with glyhb > 7 higher than the proportion of men with glyhb > 7?

```{r}
# with prop.test
diabetes$high_a1c <- diabetes$glyhb > 7
table(diabetes$high_a1c, diabetes$gender)
table(diabetes$gender[!is.na(diabetes$high_a1c)]) ## important to remove missing - otherwise proportions are different
prop.test(c(34, 26), c(228, 162))

# all in one
# using ! with diabetes$high_a1c to get the table in the expected order
prop.test(table(diabetes$gender, !diabetes$high_a1c))

# with infer package:
library(infer)
diabetes %>% 
  mutate(high_a1c = glyhb > 7) %>%
  prop_test(high_a1c ~ gender)

# equivalent to chi-squared test:
chisq.test(diabetes$high_a1c, diabetes$gender)
```



## Exercise

Make a new variable indicating whether a participant's waist measurement is greater than their hip measurement.  Is this new variable independent of gender?

```{r}
diabetes$waist_larger <- diabetes$waist > diabetes$hip
tabyl(diabetes, waist_larger, gender) 
chisq.test(diabetes$waist_larger, diabetes$gender)
```


## Exercise

Run a linear regression model with glyhb as the dependent variable and physical measurements of participants (height, weight, waist, hip) and gender as predictor variables.  Which of these variables are significant predictors?

```{r}
reg1 <- lm(glyhb ~ gender + height + weight + waist + hip, data = diabetes)
summary(reg1)
```


## Exercise

Is frame size independent of the height of participants?

What if you compute it by gender?

```{r}
# look at the values first:
diabetes %>%
  group_by(frame) %>%
  summarize(avg_height = mean(height, na.rm=TRUE)) 

diabetes %>%
  group_by(frame, gender) %>%
  summarize(avg_height = mean(height, na.rm=TRUE)) %>%
  # put in a table format:
  pivot_wider(names_from=gender, values_from=avg_height)

# how many participants in each group?
tabyl(diabetes, frame, gender)

aov(height ~ frame, data = diabetes) %>% summary()

aov(height ~ frame, data = diabetes[diabetes$gender == "male", ]) %>% summary()
aov(height ~ frame, data = diabetes[diabetes$gender == "female", ]) %>% summary()

# results above are different than for a two-way anova
aov(height ~ frame + gender, data = diabetes) %>% summary()
# because they're testing slightly different things

```


## Exercise



```{r}

```


## Exercise



```{r}

```


## Exercise



```{r}

```


## Exercise



```{r}

```


## Exercise



```{r}

```


## Exercise



```{r}

```


## Exercise



```{r}

```


