---
title: "Weeks 6 & 7 Exercises (Answers)"
format: html
project:
  execute-dir: project
editor_options: 
  chunk_output_type: console
---

```{r}
library(tidyverse)
library(janitor)
```

Load `data/diabetes.csv`.  Basic information on the dataset is available here: https://hbiostat.org/data/repo/diabetes


```{r}
diabetes <- read_csv("data/diabetes.csv")
```


## Exercise 

There are two drugs, A and B.  If A cures 15% of patients, and B cures 18% of patients.  Simulate an experiment where each drug is taken by 20 different patients by taking random draws from a binomial distribution.  The n argument is the number of experiments, and the size argument is the number of people

Which drug cured more patients in the simulated experiment?

```{r}
rbinom(n=1, size=20, prob=.15)
rbinom(n=1, size=20, prob=.18)

# your answer output may differ due to random seeds being different
```


## Exercise 

Repeat the exercise above, but run the experiment 10 times.  In how many instances does drug B cure more people than drug A?

```{r}
drugA <- rbinom(n=10, size=20, prob=.15)
drugB <- rbinom(n=10, size=20, prob=.18)

sum(drugB > drugA)

# your answer output may differ due to random seeds being different
```

## Exercise 

* Get the mean and standard deviation of the bp.1s variable in the diabetes dataset.  
* Create a normal distribution with this mean and standard deviation.  
* Then compute the 75th and 90th percentiles from both the theoretical normal distribution and from the actual data.

How do these compare?

```{r}
qnorm(c(.75, .9),
      mean=mean(diabetes$bp.1s, na.rm=TRUE),
      sd=sd(diabetes$bp.1s, na.rm=TRUE))

quantile(diabetes$bp.1s, 
         probs=c(.75, .9),
         na.rm=TRUE)
```


## Exercise 

Create a new variable chol_z that converts the chol variable into z-scores.

```{r}
diabetes$chol_z <- (diabetes$chol - mean(diabetes$chol, na.rm=TRUE)) /
  sd(diabetes$chol, na.rm=TRUE)
```


## Exercise 

Is the average height of male participants greater than that of female participants?  Use a t-test.

```{r}
t.test(diabetes$height ~ diabetes$gender)
```


## Exercise 

Is the waist to hip ratio different for male and female participants?

```{r}
diabetes$waist_hip <- diabetes$waist / diabetes$hip
t.test(diabetes$waist_hip ~ diabetes$gender)
```


## Exercise 

Is the cholesterol level of participants with glyhb > 7 higher than those with lower glyhb levels?

```{r}
diabetes$high_a1c <- diabetes$glyhb > 7
t.test(diabetes$chol ~ diabetes$high_a1c)
```


## Exercise 

Compute a 95% confidence interval on the average stab.glu value

```{r}
avg <- mean(diabetes$stab.glu, na.rm=TRUE) 
se <- sd(diabetes$stab.glu, na.rm=TRUE)/sqrt(nrow(diabetes))
avg + 1.96 * se # we have a lot of observations - using 1.96
avg - 1.96 * se
avg
```


## Exercise 

Compute a 95% confidence interval on the average stab.glu value for patients 80 years old or older.

```{r}
tval <- qt(.05/2, sum(diabetes$age >= 80) - 1)
avg <- mean(diabetes$stab.glu[diabetes$age >= 80], na.rm=TRUE) 
se <- sd(diabetes$stab.glu[diabetes$age >= 80], na.rm=TRUE)/sqrt(sum(diabetes$age >= 80))
avg + tval * se
avg - tval * se
avg
```


## Exercise 

For participants with 2 blood pressure readings, is the second reading higher than the first?

```{r}
# Order matters here!  Second value will be subtracted from first value
t.test(diabetes$bp.2s, diabetes$bp.1s, paired=TRUE)
t.test(diabetes$bp.2d, diabetes$bp.1d, paired=TRUE)

# could specify the alternative since "greater" is in the question:
t.test(diabetes$bp.2s, diabetes$bp.1s, paired=TRUE, alternative = "greater")
t.test(diabetes$bp.2d, diabetes$bp.1d, paired=TRUE, alternative = "greater")

# check that the differences are the right direction:
mean(diabetes$bp.2s - diabetes$bp.1s, na.rm=TRUE)
# second reading values are on average LOWER than first reading values
```


## Exercise 

What is the correlation between the two blood sugar measures (glyhb, stab.glu)? Is it significantly different from 0?

```{r}
cor.test(diabetes$stab.glu, diabetes$glyhb, use="pairwise")
```



## Exercise 

What percentile is a participant with a glyhb value of 12?  Hint: don't use the quantile function for this - this is the reverse question.  What proportion of participants have a value less than this?

```{r}
sum(diabetes$glyhb <= 12, na.rm=TRUE)/sum(!is.na(diabetes$glyhb))

# reverse (won't match exactly, due to how quantiles are computed):
quantile(diabetes$glyhb, .9666, na.rm=TRUE)
```


## Exercise

How many standard deviations above average is a participant with a systolic blood pressure of 170?

```{r}
(170 - mean(diabetes$bp.1s, na.rm=TRUE))/sd(diabetes$bp.1s, na.rm=TRUE)
```


## Exercise

Is the height of male participants from the locations the same?

```{r}
diabetes %>%
  filter(gender == "male") %>%
  group_by(location) %>%
  summarize(mean(height))

t.test(height ~ location, data = diabetes[diabetes$gender == "male",])

```


## Exercise 

Is the proportion of women in the data with glyhb > 7 higher than the proportion of men with glyhb > 7?

```{r}
# with prop.test
diabetes$high_a1c <- diabetes$glyhb > 7
table(diabetes$high_a1c, diabetes$gender)
table(diabetes$gender[!is.na(diabetes$high_a1c)]) ## important to remove missing - otherwise proportions are different
prop.test(c(34, 26), c(228, 162))

# all in one
# using ! with diabetes$high_a1c to get the table in the expected order
prop.test(table(diabetes$gender, !diabetes$high_a1c))

# with infer package:
library(infer)
diabetes %>% 
  mutate(high_a1c = glyhb > 7) %>%
  prop_test(high_a1c ~ gender)

# equivalent to chi-squared test:
chisq.test(diabetes$high_a1c, diabetes$gender)
```

## Exercise

Make a new variable indicating whether a participant's waist measurement is greater than their hip measurement.  Is this new variable independent of gender?

```{r}
diabetes$waist_larger <- diabetes$waist > diabetes$hip
tabyl(diabetes, waist_larger, gender) 
chisq.test(diabetes$waist_larger, diabetes$gender)
```


## Exercise

Run a linear regression model with glyhb as the dependent variable and physical measurements of participants (height, weight, waist, hip) and gender as predictor variables.  Which of these variables are significant predictors?

```{r}
reg1 <- lm(glyhb ~ gender + height + weight + waist + hip, data = diabetes)
summary(reg1)
```





## Challenge Exercise 

Which of the variables in the diabetes data are notably skewed?  Meaning that the distribution is not symmetrical around the mean - based on human estimation, not a statistical test.  Look at stab.glu, glyhb, chol, hdl, bp.1s, bp.1d

There are a few different ways you could investigate this - computing summary measures and plotting can both be useful.  If you're plotting, the boxplot() function might be useful.

```{r}
boxplot(diabetes$chol)
hist(diabetes$chol)

# The following would generally be considered skewed: stab.glu, glyhb, hdl 
# the following aren't very skewed: chol, bp.1d, bp.1s (this one is a little off because it's truncated at 0, but it's still roughly normal)
```




