---
title: "Week 5 Exercises (Answers)"
format: html
project:
  execute-dir: project
editor_options: 
  chunk_output_type: console
---


```{r}
library(tidyverse)
library(janitor)
```

Load `data/diabetes.csv`.  Basic information on the dataset is available here: https://hbiostat.org/data/repo/diabetes


```{r}
diabetes <- read_csv("data/diabetes.csv")
```


## Exercise 

Convert the gender variable to a factor.

```{r}
diabetes$gender <- factor(diabetes$gender)

# or
diabetes <- mutate(diabetes, gender = factor(gender))
```


## Exercise 

Convert the frame variable to a factor.  Set the levels in the appropriate order.

Do not make the factor ordered (because we're going to do other things with it later).

```{r}
diabetes$frame <- factor(diabetes$frame, levels=c("small", "medium", "large"))

# or 
diabetes <- mutate(diabetes, frame = factor(frame, levels=c("small", "medium", "large")))
```


## Exercise 

Are there any missing values in the frame variable?

```{r}
sum(is.na(diabetes$frame))  # yes
```


## Exercise 

Make missing (NA) an explicit level in the frame factor variable.  Relabel it as "unknown".

```{r}
diabetes$frame <- addNA(diabetes$frame)
levels(diabetes$frame)[4] <- "unknown"

# or
diabetes <- mutate(diabetes, frame = addNA(frame))
levels(diabetes$frame)[4] <- "unknown"  # there's not an easy way to do this within dplyr
```


## Exercise 

How many observations are missing a weight?  How many are missing a height?  Are there any observations missing both?

```{r}
sum(is.na(diabetes$weight))
sum(is.na(diabetes$height))

sum(is.na(diabetes$weight) & is.na(diabetes$height)) # none missing both
```


## Exercise 

How many observations in the data are complete (no missing values)?  

What proportion of observations are complete? (hint: you need two pieces of information to compute this - you know how to get both, then combine them)

```{r}
# counting complete cases
sum(complete.cases(diabetes))
# or
dim(diabetes[complete.cases(diabetes), ])

# counting total observations
nrow(diabetes)
# or 
length(complete.cases(diabetes))

# combine them
sum(complete.cases(diabetes))/nrow(diabetes)
```


## Exercise 

Which column has the most missing values?

```{r}
diabetes %>%
  summarize(across(everything(), \(x) sum(is.na(x))))

# or
summary(diabetes)
```


## Exercise 

What's the correlation between weight and height?

```{r}
cor(diabetes$weight, diabetes$height, use="pairwise")
```


## Exercise 

What's the correlation between weight and height by frame size?

```{r}
diabetes %>%
  group_by(frame) %>%
  summarize(cor(weight, height, use="pairwise"))
```


## Exercise 

Compute a correlation matrix for the variables height, weight, waist, and hip

```{r}
cor(diabetes[,c("height", "weight", "waist", "hip")], use="pairwise")
```


## Challenge Exercise 

You should see a surprising correlation in the answer above.  Is there another variable that might help explain that?

```{r}
diabetes %>%
  group_by(gender) %>%
  summarize(cor(hip, height, use="pairwise"))
```


## Exercise 

What's the correlation between stab.glu (instant blood sugar level) and glyhb (A1C)?

```{r}
cor(diabetes$stab.glu, diabetes$glyhb, use="pairwise")
```


## Exercise 

Plot stab.glu (instant blood sugar level) and glyhb (A1C).  

```{r}
plot(diabetes$stab.glu, diabetes$glyhb)
```


## Challenge Exercise 

Do you see any unexpected values in the plot above?  What are the study ID numbers of two observations that stand out?

```{r}
diabetes$id[diabetes$stab.glu < 150 & diabetes$glyhb > 14]
diabetes$id[diabetes$stab.glu > 350 & diabetes$glyhb < 6]
```


## Exercise 

For the patients with all 4 blood pressure readings available (not missing), what is the correlation between the first and second readings?  

```{r}
# use cor() missing options
cor(diabetes[, c("bp.1s", "bp.1d", "bp.2s", "bp.2d")], use="complete")

# another option
bps <- diabetes[, c("bp.1s", "bp.1d", "bp.2s", "bp.2d")]
all_bp <- bps[complete.cases(bps), ]
cor(all_bp)
# or
cor(all_bp$bp.1s, all_bp$bp.2s)
cor(all_bp$bp.1d, all_bp$bp.2d)
```


## Exercise 

Do older participants have higher blood pressure readings?

```{r}
# with correlation
cor(diabetes$age, diabetes$bp.1s, use="pairwise")
cor(diabetes$age, diabetes$bp.1d, use="pairwise")
```


## Exercise

Part 1: How many participants are over 60?

Part 2: Do those participants have higher blood pressures than younger participants on average?

```{r}
sum(diabetes$age > 60)

diabetes %>%
  mutate(over60 = age > 60) %>%
  group_by(over60) %>%
  summarize(mean(bp.1s, na.rm=TRUE),
            mean(bp.2s, na.rm=TRUE))
```


## Challenge Exercise

How do the LDL and HDL cholesterol levels of those who have a high blood sugar reading (on either measure) compare to those who don't have a high blood sugar reading?  Cholesterol = LDL + HDL

```{r}
diabetes %>%
  mutate(high_sugar = glyhb > 7 | stab.glu > 120) %>%
  group_by(high_sugar) %>%
  summarize(mean(chol-hdl, na.rm=TRUE),
            mean(hdl, na.rm=TRUE))
```



## Challenge Exercise

Is BMI (703 * weight รท (height ^ 2)) or the waist-hip ratio a better predictor of blood sugar?  Is either a very good predictor?  Plotting the data may help you assess this.

```{r}
diabetes <- diabetes %>%
  mutate(bmi = 703*weight / (height^2),
         waist_hip = waist/hip) 

diabetes %>%
  summarize(cor(bmi, glyhb, use="pair"),
            cor(bmi, stab.glu, use="pair"),
            cor(waist_hip, glyhb, use="pair"),
            cor(waist_hip, stab.glu, use="pair"))

plot(diabetes$bmi, diabetes$stab.glu)
plot(diabetes$bmi, diabetes$glyhb)
plot(diabetes$waist_hip, diabetes$stab.glu)
plot(diabetes$waist_hip, diabetes$glyhb)
plot(diabetes$bmi, diabetes$waist_hip)
```


