---
title: "Week 7 Exercises (Answers)"
format: html
project:
  execute-dir: project
editor_options: 
  chunk_output_type: console
---


```{r}
library(tidyverse)
library(janitor)
```

Load `data/diabetes.csv`.  Basic information on the dataset is available here: https://hbiostat.org/data/repo/diabetes


```{r}
diabetes <- read_csv("data/diabetes.csv")
```


## Exercise 

Is the cholesterol level of participants with glyhb > 7 higher than those with lower glyhb levels?

```{r}


```



## Exercise 

Is the average height of male participants greater than that of female participants?  Use a t-test.

```{r}


```


## Exercise 

Is the waist to hip ratio different for male and female participants?

```{r}


```


## Exercise 

For participants with 2 blood pressure readings, is the second reading higher than the first?

```{r}


```


## Exercise

Is the height of male participants from the locations the same?

```{r}


```





## Exercise 

Is the proportion of women in the data with glyhb > 7 higher than the proportion of men with glyhb > 7?

```{r}


```



## Exercise

Make a new variable indicating whether a participant's waist measurement is greater than their hip measurement.  Is this new variable independent of gender?

```{r}


```


## Exercise

Is frame size independent of the height of participants?

What if you compute it by gender?

```{r}


```



## Exercise

Run a linear regression model with glyhb as the dependent variable and physical measurements of participants (height, weight, waist, hip) and gender as predictor variables.  Which of these variables are significant predictors?

```{r}


```


