---
title: "Week 6 Exercises (Answers)"
format: html
project:
  execute-dir: project
editor_options: 
  chunk_output_type: console
---


```{r}
library(tidyverse)
library(janitor)
```

Load `data/diabetes.csv`.  Basic information on the dataset is available here: https://hbiostat.org/data/repo/diabetes


```{r}
diabetes <- read_csv("data/diabetes.csv")
```


## Exercise 

There are two drugs, A and B.  If A cures 15% of patients, and B cures 18% of patients.  Simulate an experiments where each drug is taken by 20 different patients.  Which drug cured more patients in the experiment?

```{r}


```


## Exercise 

Repeat the exercise above, but run the experiment 10 times.  In how many instances does drug B cure more people than drug A?

```{r}


```


## Exercise 

* Get the mean and standard deviation of the bp.1s variable in the diabetes dataset.  
* Create a normal distribution with this mean and standard deviation.  
* Then compute the 75th and 90th percentiles from both the theoretical normal distribution and from the actual data.

How do these compare?

```{r}


```


## Exercise 

Create a new variable chol_z that converts the chol variable into z-scores.

```{r}


```


## Exercise 

Is the average height of male participants greater than that of female participants?  Use a t-test.

```{r}


```


## Exercise 

Is the waist to hip ratio different for male and female participants?

```{r}


```



## Exercise 

Is the cholesterol level of participants with glyhb > 7 higher than those with lower glyhb levels?

```{r}


```


## Exercise 

Compute a 95% confidence interval on the average stab.glu value

```{r}


```


## Exercise 

Compute a 95% confidence interval on the average stab.glu value for patients 80 years old or older.

```{r}


```


## Exercise 

For participants with 2 blood pressure readings, is the second reading higher than the first?

```{r}


```


## Exercise 

Create two random samples of 20 participants each from the diabetes data.  Is the mean stab.glu value different for the two samples?

```{r}


```


## Exercise 

What is the correlation between the two blood sugar measures (glyhb, stab.glu)? Is it significantly different from 0?

```{r}


```



## Exercise 

What percentile is a participant with a glyhb value of 12?

```{r}


```


## Exercise

How many standard deviations above average is a participant with a systolic blood pressure of 170?

```{r}


```


## Exercise

Is the height of male participants from the locations the same?

```{r}


```



## Challenge Exercise 

Which of the variables in the diabetes data are notably skewed?  Meaning that the distribution is not symmetrical around the mean - based on human estimation, not a statistical test.  Look at stab.glu, glyhb, chol, hdl, bp.1s, bp.1d

There are a few different ways you could investigate this - computing summary measures and plotting can both be useful.  If you're plotting, the boxplot() function might be useful.

```{r}


```



