---
title: "Week 7 Exercises (Answers)"
format: html
project:
  execute-dir: project
editor_options: 
  chunk_output_type: console
---


```{r}
library(tidyverse)
library(janitor)
```

Load `data/diabetes.csv`.  Basic information on the dataset is available here: https://hbiostat.org/data/repo/diabetes


```{r}
diabetes <- read_csv("data/diabetes.csv")
```


## Exercise 

Is the proportion of women in the data with glyhb > 7 higher than the proportion of men with glyhb > 7?

```{r}


```



## Exercise

Make a new variable indicating whether a participant's waist measurement is greater than their hip measurement.  Is this new variable independent of gender?

```{r}


```


## Exercise

Run a linear regression model with glyhb as the dependent variable and physical measurements of participants (height, weight, waist, hip) and gender as predictor variables.  Which of these variables are significant predictors?

```{r}


```


## Exercise

Is frame size independent of the height of participants?

What if you compute it by gender?

```{r}


```


More coming soon...

## Exercise



```{r}

```


## Exercise



```{r}

```


## Exercise



```{r}

```


## Exercise



```{r}

```


## Exercise



```{r}

```


## Exercise



```{r}

```


## Exercise



```{r}

```


