---
title: "Week 5 Exercises (Answers)"
format: html
project:
  execute-dir: project
editor_options: 
  chunk_output_type: console
---


```{r}
library(tidyverse)
library(janitor)
```

Load `data/diabetes.csv`.  Basic information on the dataset is available here: https://hbiostat.org/data/repo/diabetes


```{r}
diabetes <- read_csv("data/diabetes.csv")
```


## Exercise 

Convert the gender variable to a factor.

```{r}


```


## Exercise 

Convert the frame variable to a factor.  Set the levels in the appropriate order.

Do not make the factor ordered (because we're going to do other things with it later).

```{r}


```


## Exercise 

Are there any missing values in the frame variable?

```{r}


```


## Exercise 

Make missing (NA) an explicit level in the frame factor variable.  Relabel it as "unknown".

```{r}


```


## Exercise 

How many observations are missing a weight?  How many are missing a height?  Are there any observations missing both?

```{r}


```


## Exercise 

How many observations in the data are complete (no missing values)?  

What proportion of observations are complete? (hint: you need two pieces of information to compute this - you know how to get both, then combine them)

```{r}


```


## Exercise 

Which column has the most missing values?

```{r}


```


## Exercise 

What's the correlation between weight and height?

```{r}


```


## Exercise 

What's the correlation between weight and height by frame size?

```{r}


```


## Exercise 

Compute a correlation matrix for the variables height, weight, waist, and hip

```{r}


```


## Challenge Exercise 

You should see a surprising correlation in the answer above.  Is there another variable that might help explain that?

```{r}


```


## Exercise 

What's the correlation between stab.glu (instant blood sugar level) and glyhb (A1C)?

```{r}


```


## Exercise 

Plot stab.glu (instant blood sugar level) and glyhb (A1C).  

```{r}


```


## Challenge Exercise 

Do you see any unexpected values in the plot above?  What are the study ID numbers of two observations that stand out?

```{r}


```


## Exercise 

For the patients with all 4 blood pressure readings available (not missing), what is the correlation between the first and second readings?  

```{r}


```


## Exercise 

Do older participants have higher blood pressure readings?

```{r}


```


## Exercise

Part 1: How many participants are over 60?

Part 2: Do those participants have higher blood pressures than younger participants on average?

```{r}


```


## Challenge Exercise

How do the LDL and HDL cholesterol levels of those who have a high blood sugar reading (on either measure) compare to those who don't have a high blood sugar reading?  Cholesterol = LDL + HDL

```{r}


```



## Challenge Exercise

Is BMI (703 * weight รท (height ^ 2)) or the waist-hip ratio a better predictor of blood sugar?  Is either a very good predictor?  Plotting the data may help you assess this.

```{r}


```


## Exercise

Use ifelse() to make a new variable bmi_cat that codes "<= 30" and "30+" categories.  Make a table of the new variable - make bmi_cat a factor so the categories are in an appropriate order.

Challenge: Make the BMI category variable with 3 categories by nesting calls to the ifelse function.  

```{r}


```


## Exercise

How much more likely are participants with glyhb > 7 to have a stab.glu > 100 than those with glyhb <= 7?  (Compute the odds ratio)

```{r}


```




